---
title: "Digital Media Use and Trust in Institutions"
subtitle: "A data cleaning & descriptive analysis walkthrough in R"
author: "Raya Sengupta"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

## About this project

This project demonstrates a standard applied workflow in survey-based social
research: taking a raw, messy data export and turning it into a clean data set
ready for descriptive analysis.

**Note on the data:** the survey responses used here are simulated, but the
questionnaire items are modeled directly on real survey instruments, the
European Social Survey's media-use module and the standard 0–10 institutional
trust battery used in the ESS and World Values Survey. Real cross-national
survey exports (raw `.csv`/`.sav` files) commonly contain exactly the kinds
of problems handled below: inconsistent category labels, numeric missing-value
codes (e.g. `98`/`99` for "don't know"/"refused"), blank strings, stray
whitespace, and duplicate rows from export errors. Simulating this lets the
cleaning steps below reflect real research practice while keeping the project
self-contained and reproducible for anyone reviewing it.

The research question this mirrors: **does frequency of online news/social
media use relate to how much people trust core institutions (parliament,
press, science, politicians)?** a live question in the digital
disinformation/media-literacy literature.

## 1. Load the raw data

```{r load}
raw <- read.csv("raw_survey_export.csv", stringsAsFactors = FALSE)
str(raw)
```

```{r glimpse}
head(raw, 5)
```

The raw export has `r nrow(raw)` rows and `r ncol(raw)` columns — already more
rows than expected, a first sign of duplicates to check for.

## 2. Clean the data

### 2.1 Remove duplicate and incomplete rows

```{r dedupe}
clean <- raw %>%
  distinct(resp_id, .keep_all = TRUE) %>%     # drop duplicate respondents
  filter(!is.na(country))                     # drop rows with no country recorded

cat("Rows after de-duplication:", nrow(clean), "\n")
```

### 2.2 Standardise the education variable

Raw category labels are inconsistent (`"Secondary"` vs `"secondary "` vs
trailing whitespace vs `"NA"` as a literal string).

```{r clean_education}
clean <- clean %>%
  mutate(
    education = trimws(tolower(education)),
    education = na_if(education, "na"),
    education = na_if(education, ""),
    education = recode(education,
                        "phd" = "tertiary"),   # fold PhD into tertiary for this analysis
    education = factor(education,
                        levels = c("primary", "secondary", "tertiary"),
                        ordered = TRUE)
  )

table(clean$education, useNA = "ifany")
```

### 2.3 Recode internet/social media news use into an ordered factor

```{r clean_netuse}
clean <- clean %>%
  mutate(
    net_use_news = na_if(net_use_news, "98"),
    net_use_news = na_if(net_use_news, "99"),
    net_use_news = factor(net_use_news,
      levels = c("Never", "Less than once a month", "Once a month",
                 "Several times a month", "Once a week",
                 "Several times a week", "Every day"),
      ordered = TRUE)
  )

table(clean$net_use_news, useNA = "ifany")
```

### 2.4 Clean the trust items (0–10 scales)

Survey exports typically use out-of-range numeric codes for missing data
(here `98` = "don't know", `99` = "refused"); these need to become proper
`NA`s before any averaging, or they will silently wreck the descriptive
statistics.

```{r clean_trust}
trust_vars <- c("trust_parliament", "trust_press", "trust_science", "trust_politicians")

clean <- clean %>%
  mutate(across(all_of(trust_vars), as.character)) %>%
  mutate(across(all_of(trust_vars), ~ na_if(.x, ""))) %>%
  mutate(across(all_of(trust_vars), ~ suppressWarnings(as.numeric(.x)))) %>%
  mutate(across(all_of(trust_vars), ~ ifelse(.x %in% c(98, 99), NA, .x)))

summary(clean[trust_vars])
```

### 2.5 Build a composite trust index

A common step in this literature: average the four trust items into a single
institutional trust index per respondent.

```{r trust_index}
clean <- clean %>%
  rowwise() %>%
  mutate(trust_index = mean(c_across(all_of(trust_vars)), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(trust_index = ifelse(is.nan(trust_index), NA, trust_index))

summary(clean$trust_index)
```

## 3. Descriptive analysis

### 3.1 Sample composition

```{r sample_comp}
clean %>%
  count(country) %>%
  arrange(desc(n))
```

```{r edu_comp}
clean %>%
  filter(!is.na(education)) %>%
  count(education) %>%
  mutate(pct = percent(n / sum(n)))
```

### 3.2 Distribution of the trust index

```{r trust_dist_plot}
ggplot(clean %>% filter(!is.na(trust_index)), aes(x = trust_index)) +
  geom_histogram(binwidth = 0.5, fill = "#2c6e91", color = "white") +
  labs(title = "Distribution of institutional trust index",
       subtitle = "Average of trust in parliament, press, science, and politicians (0-10 scale)",
       x = "Trust index (0-10)", y = "Number of respondents") +
  theme_minimal(base_size = 12)
```

### 3.3 Trust by frequency of online news/social media use

This is the key descriptive relationship: mean trust index by how often
respondents use the internet/social media for news.

```{r trust_by_netuse}
by_netuse <- clean %>%
  filter(!is.na(net_use_news), !is.na(trust_index)) %>%
  group_by(net_use_news) %>%
  summarise(
    mean_trust = mean(trust_index, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

by_netuse
```

```{r trust_by_netuse_plot, fig.width=8}
ggplot(by_netuse, aes(x = net_use_news, y = mean_trust)) +
  geom_col(fill = "#c15b3c") +
  labs(title = "Mean institutional trust by frequency of online news use",
       x = "Frequency of using internet/social media for news",
       y = "Mean trust index (0-10)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

### 3.4 Trust by education level

```{r trust_by_edu}
clean %>%
  filter(!is.na(education), !is.na(trust_index)) %>%
  group_by(education) %>%
  summarise(mean_trust = mean(trust_index), sd_trust = sd(trust_index), n = n(),
            .groups = "drop")
```

## 4. Summary

Starting from a raw export with duplicate rows, inconsistent category
labels, and numeric missing-value codes, this workflow produced:

- a de-duplicated, cleanly typed dataset (`r nrow(clean)` respondents)
- a validated composite institutional trust index
- descriptive breakdowns of trust by online news-use frequency and by
  education level, with accompanying visualizations



```{r save_clean, include=FALSE}
write.csv(clean, "clean_survey_data.csv", row.names = FALSE)
```
